---
title: "estimated marginal means"
author: "luke reding"
date: "3/10/2017"
output: html_document
---

The goal here is to show how to use the `lsmeans` package in R to get the estimated marginal means from a model.

## What are estimated marginal means?

You have a model, but it's complex. There might be a lot of covariates, or you might have a random effect or two. In these cases it's difficult to simply show the data to show an effect. Instead, you can use estimated marginal means, or covariate-adjusted means. This allows you to show some effect you're interested in independent of the effect that other varaibles in your model have.

## example using a linear model (lm())

Let's use the `oranges` dataset that comes with `lsmeans` to flesh this out.

```{r}
library("lsmeans")
oranges.lm1 <- lm(sales1 ~ price1 + price2 + day + store, data = oranges)
car::Anova(oranges.lm1)

```

Part of what happens with the `lsmeans` package is that a _reference grid_ is established. A reference grid is all the combinations of the factor levels of all the variables in the analysis; any numeric variables only get one level: their mean.

To see what this grid looks like for the `oranges` dataset:


```{r}

( oranges.rg1 <- ref.grid(oranges.lm1) )
summary(oranges.rg1)


```


The ANOVA showed a significant effect of _day_. Let's compare plotting the actual data to extracting and plotting the estimated marginal means:

```{r}

library(tidyverse)
library(magrittr)
source("~/Documents/random_scripts/plotting_functions.R")

ggplot(oranges, aes(y = sales1, x = day)) + 
  geom_point() + 
  theme_mod() +
  add_axes()

```

We can get the estimated marginal means by using:


```{r}

(marginal <- lsmeans(oranges.lm1, "day") %>% broom::tidy(.))


```

This is the estimated `sales1` for each day averaged over all the variables in our model, `oranges.lm1`. I've also used the `broom` package here to make the output more useable and informatibe.


We could now plot these estimated means instead of the raw data:


```{r}
marginal %>%
  ggplot(aes(x = day, y = estimate)) +
  geom_col() +
  geom_errorbar(aes(ymin=estimate-std.error, ymax=estimate+std.error), width = 0) +
  theme_mod()

## or
marginal %>%
  ggplot(aes(x = day, y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin=estimate-std.error, ymax=estimate+std.error), width = 0) +
  theme_mod()

```

Here the effect is a bit more clearly seen and is more honest to the underlying statistical analysis used to analyze the data.

## more complicated

We can also do all sorts of stats on these marginal means. For example, in the previous example, we might want to test whether the price estimate on days 1 and 2 is different from the prices on the other days:


```{r}

days.lsm <- lsmeans(oranges.lm1, "day")
( days_contr.lsm <- contrast(days.lsm, "trt.vs.ctrl", ref = c(1,2)) )

```
As we can see from the plot above, day 4 and, to a lesster degree, day 6 aren't too different from the prices on days 1 and 2, and the results of the code above verify this intuition. 

You can also do something more general and compare all lsmeans to each other.

```{r}

pairs(lsmeans(oranges.lm1, "day"))

```

Again, these p-values verify our intuition about the plot above.

It also might be useful to add letters or numbers to groups that don't differ significantly:


```{r}

require(multcompView)
cld(days.lsm, alpha = .10)

```

## example using a linear mixed model (lmer())


```{r}

require("lme4")
data("Oats", package = "nlme")

Oats.lmer2 <- lmer(log(yield) ~ Variety + poly(nitro,2) + (1|Block/Variety),  data = Oats)
anova(Oats.lmer2)

ggplot(Oats, aes(x = nitro, y = log(yield), color = Variety)) +
  geom_jitter(height = 0, width = 0.03) +
  scale_color_classic() +
  theme_mod() + 
  add_axes()

## or 

Oats %>%
  group_by(Variety, nitro) %>%
  summarise(avg = mean(yield, na.rm = T)) %>%
  ggplot(aes(factor(nitro), avg, color = Variety, group = Variety)) +
  geom_line(position = position_dodge(0.3)) +
  geom_point(position = position_dodge(0.3)) +
  scale_color_classic() +
  theme_mod() +
  add_axes()

```
Again, if we want to visualize the average effect of nitrogen on yeild, averaging across varieties and blocks, we can use:


```{r}
(marginal <- lsmeans(Oats.lmer2, "nitro", cov.reduce = FALSE) %>% broom::tidy(.)) %>%
  ggplot(aes(x = factor(nitro), y = estimate)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0) +
  geom_col() +
  theme_mod()
  
## or 
marginal %>%
  ggplot(aes(x = factor(nitro), y = estimate)) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0) +
  geom_point() +
  theme_mod() +
  add_axes()
```

Or, to show the average yeild by variety averaged across blocks:

```{r}
## or 
lsmeans(Oats.lmer2, "nitro", by = "Variety", cov.reduce = FALSE) %>%
  broom::tidy(.) %>%
  ggplot(aes(nitro, estimate, color = Variety)) +
  geom_line(position = position_dodge(0.1)) +
  geom_point(position = position_dodge(0.1)) +
  geom_errorbar(width = 0, aes(ymin = conf.low, ymax = conf.high), position = position_dodge(0.1)) +
  scale_color_classic() +
  theme_mod()


```


Nitro is a number, but it should really be treated as a factor. We can use `cov.reduce = FALSE` to tell lsmeans to treat numeric variables as factors:


```{r}

(Oats.lsm2 <- lsmeans(Oats.lmer2, ~ nitro | Variety, cov.reduce = FALSE))

```



## more info

The vignette for lsmeans (`vignette("using-lsmeans")`) is super useful and a great resource.
 





