---
title: "R Notebook"
output:
  html_notebook:
    fig_height: 8
    fig_width: 8
    highlight: tango
    theme: journal
    toc: yes
  html_document:
    toc: yes
---

Very basic attempt to figure out, given a dataframe with a lot of variables, how to find potentially significant relationships between pairs of variables.

I first use the `wakefeild` package to generate some simulated data. I also load the `tidyverse` (contains `ggplot2`, `dplyr`, etc.) and `magrittr` packages:

```{r}
library(wakefield)
library(tidyverse)
library(magrittr)

# generate dataframe with fake data
set.seed(11)
df <- r_data_frame(
    n = 100,
    id,
    Scoring = rnorm,
    race,
    age,
    sex,
    iq,
    height,
    dice,
    hair
)

df %>% sample_n(10)
  
```

Actually, to keep things simple, let's only take the numeric variables:

```{r}

df %<>% select_if(is.numeric)

# check to make sure
map(df, is.numeric)

```

Let's create a new dataframe where we calculate the correlation between each pair of variables. 

Note that I'm using two calls to `map` here which is like use `apply` instead of nested `for` loops. I'm also using `broom::glance` which takes a model output and creates a dataframe out of it:

```{r}
library(broom)

# Function that returns a dataframe of the correlation test between xth and all subsequent columns of dataframe `df`
get_stats <- function(x){
  (x+1):ncol(df) %>% 
    map(~ cor.test(pull(df[,x]), pull(df[,.x]), use="pairwise.complete.obs") %>%
    glance %>%
    mutate(variable_1 = names(df[,x]),
           variable_2 = names(df[,.x]))) %>%
      reduce(bind_rows)
}

# make results dataframe that contains the results of all the `cor.test`s
results <- 1:(ncol(df)-1) %>% 
  map(get_stats) %>% 
  reduce(bind_rows)


# check to make sure there are 'ncol(df) choose 2' rows in `results`
choose(ncol(df), 2) == nrow(results)

```

Sort the results by the correlation coefficient:

```{r}

results %>% 
  arrange(desc(estimate))

```

Better, plot it:

```{r}
require(ggpubr)

results %>%
  unite(combo, variable_1, variable_2, sep = "*") %>%
  ggplot(aes(x = reorder(combo, desc(estimate)), y = estimate)) +
    geom_point(aes(size = p.value)) +
  scale_size(trans = "log10", breaks = c(0.01, 0.1, 1)) +
  theme_pubr() +
  labs(x = "trait combination", y = "correlation") +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = 2)

```